---
title: "Project3"
author: "Leithe, Martinussen & Saghagen"
date: "November, 2017"
output: #3rd letter intentation hierarchy
#  beamer_presentation:
###    incremental: true # or >* for one at a time
#  slidy_presentation:
#    font_adjustment: +1  
  prettydoc::html_pretty:
    theme: architect
    highlight: github
#   pdf_document:
#    toc: true
#    toc_depth: 2
#    engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,tidy = TRUE, message = FALSE, warning = FALSE)
library(ggplot2, quietly = TRUE, warn.conflicts = FALSE)
library(faraway, quietly = TRUE, warn.conflicts = FALSE)
library(GGally, quietly = TRUE, warn.conflicts = FALSE)
library(devtools, quietly = TRUE, warn.conflicts = FALSE)
library(MASS, quietly = TRUE, warn.conflicts = FALSE)
library(sjPlot, quietly = TRUE, warn.conflicts = FALSE)
library(ggpubr, quietly = TRUE, warn.conflicts = FALSE)
library(lme4,quietly = TRUE, warn.conflicts = FALSE)
```


```{r, quietly=TRUE}
data(jsp, package = "faraway")
sjsp <- jsp[jsp$year == 0, ]
sjsp <- sjsp[, -c(6, 7, 9)]  # removing class, id, english and year as they are not used
school= sjsp$school
class = sjsp$class
gender = sjsp$gender
social = sjsp$social
raven = sjsp$raven
math = sjsp$math
```


## a)
`ggpairs` schatter plot of the data grouped using `gender` and the covariates `social`, `raven` and `math` included. 
````{r}
ggpairs(sjsp, columns=c(4,5,6), mapping=aes(col=gender),legend=1)
````
We fit a linear model with `math` as response, and `raven` and `gender` as covariates
````{r}
fit = lm(math~raven+gender)
fit$coefficients
````

## b)
Measurement model for school $i$:
$$
\mathbf{Y}_i = \mathbf{X}_i \beta + \gamma_{0i} + \varepsilon_i
$$

* $\mathbf{Y}_i: n_i \times 1$ random vector of responses
* $\mathbf{X}_i: n_i \times p$ design matrix
* $\gamma_{0i}$: deviation from the population intercept, random variable
* $\varepsilon_i:$

Distributional assumptions for $\gamma_{0i}$ and $\varepsilon_i$:
$$
\gamma_{0i} \sim N(0,\tau^2_0) \quad \varepsilon \sim N(0, \sigma^2)
$$
Observations from different schools are independent. 
````{r}
fitRIa <- lmer(math ~ raven + gender + (1 | school), data = sjsp)
summary(fitRIa)
````

````{r,echo=FALSE}
coefs = data.frame(coef(summary(fitRIa)))
confinterval = confint(fitRIa)
````

````{r}
pvalue_raven <- 2 * pnorm(abs(coefs$t.value[3]), lower.tail = FALSE)
pvalue_rave
femaleconf = confinterval[5,]
femaleconf
````

## c)
$$
Corr(Y_{ij},Y_{il})=\frac{Cov(Y_{ij},Y_{il})}{\sqrt{Var(Y_{ij}Var_{il})}} = \frac{\tau^2_0}{\tau^2_0+\sigma^2}\: \text{for} j\neq i
$$
```{r}
fitRI <- lmer(math ~ raven + (1 | school), data = sjsp)
summary(fitRI)
```

```{r, echo=FALSE}
randomvar = as.data.frame(VarCorr(fitRI))[,4]
```

```{r}
corr = randomvar[1]/(randomvar[1]+randomvar[2])
corr
```
Predicted value for the random intercept:
$$
\hat{\gamma_{0i}} = \mathbf{\hat{Q}}\mathbf{U}^T_i\mathbf{\hat{V}}^{-1}_i(\mathbf{Y}_i-\mathbf{X}_i\hat{\beta})
$$

* $U_i: n_i \times (q+1)$ is the design matrix for random effects
* $V_i = Cov(U_i\gamma_i)+Cov(\varepsilon_i)$

## d)
```{r}
fitRIb <- lmer(math ~ raven + social + (1 | school), data = sjsp)
anova(fitRI, fitRIb)
```

## e)
